NDNS multilevel LCA CW2CB2 or CW2CB3

### 2018-08-18 paper/Appendices/AppendixD.tex
appendix done
limitation 1

\chapter{R \& Stata codes for processing individual level data analysis} % Main appendix title

@@

\CommentTok{# read the individual level data sets ----------------------------------------}

\NormalTok{blood78 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr7-8a_indiv.dta"}\NormalTok{)}
\NormalTok{blood56 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr5-6a_indiv.dta"}\NormalTok{)}
\NormalTok{blood14 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr1-4a_indiv_uk.dta"}\NormalTok{)}
\NormalTok{food14 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr1-4a_personleveldietarydata_uk.dta"}\NormalTok{)}
\NormalTok{food56 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr5-6a_personleveldietarydata.dta"}\NormalTok{)}
\NormalTok{food78 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr7-8a_personleveldietarydata.dta"}\NormalTok{)}
\KeywordTok{names}\NormalTok{(blood78)[}\KeywordTok{names}\NormalTok{(blood78)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}
\KeywordTok{names}\NormalTok{(blood56)[}\KeywordTok{names}\NormalTok{(blood56)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}
\KeywordTok{names}\NormalTok{(blood14)[}\KeywordTok{names}\NormalTok{(blood14)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}
\KeywordTok{names}\NormalTok{(food78)[}\KeywordTok{names}\NormalTok{(food78)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}
\KeywordTok{names}\NormalTok{(food56)[}\KeywordTok{names}\NormalTok{(food56)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}
\KeywordTok{names}\NormalTok{(food14)[}\KeywordTok{names}\NormalTok{(food14)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}

@@

\NormalTok{blood14 <-}\StringTok{ }\NormalTok{blood14 }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(ID, Ndays)}

\NormalTok{blood56 <-}\StringTok{ }\NormalTok{blood56 }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(ID, Ndays)}

\NormalTok{blood78 <-}\StringTok{ }\NormalTok{blood78 }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(ID, NDays) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{rename}\NormalTok{(}\DataTypeTok{Ndays =}\NormalTok{ NDays)}

\NormalTok{NDAYS <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(blood14, blood56, blood78)}
\NormalTok{NDAYS}\OperatorTok{$}\NormalTok{ID <-}\StringTok{ }\KeywordTok{as.numeric}\NormalTok{(NDAYS}\OperatorTok{$}\NormalTok{ID)}

@@

\NormalTok{## dta ready to be analysed in STATA}
 \CommentTok{# change the path accordingly}
\KeywordTok{write_dta}\NormalTok{(CW3CB3_regss, }\StringTok{"../CW3CB3_7regss.dta"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

**********************************************************
// variables need to be log transfomred                 //
**********************************************************

>>>>

\chapter{R and STATA codes for processing individual level data analysis} % Main appendix title

@@

\CommentTok{# read the individual level data sets ----------------------------------------}
\CommentTok{# Change the data path accordingly ----------------------------------------}
\KeywordTok{setwd}\NormalTok{(}\StringTok{".../UKDA-6533-stata11_se/stata11_se/"}\NormalTok{) }
\NormalTok{blood78 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr7-8a_indiv.dta"}\NormalTok{)}
\NormalTok{blood56 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr5-6a_indiv.dta"}\NormalTok{)}
\NormalTok{blood14 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr1-4a_indiv_uk.dta"}\NormalTok{)}
\NormalTok{food14 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr1-4a_personleveldietarydata_uk.dta"}\NormalTok{)}
\NormalTok{food56 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr5-6a_personleveldietarydata.dta"}\NormalTok{)}
\NormalTok{food78 <-}\StringTok{ }\KeywordTok{read_dta}\NormalTok{(}\StringTok{"ndns_rp_yr7-8a_personleveldietarydata.dta"}\NormalTok{)}
\KeywordTok{names}\NormalTok{(blood78)[}\KeywordTok{names}\NormalTok{(blood78)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}
\KeywordTok{names}\NormalTok{(blood56)[}\KeywordTok{names}\NormalTok{(blood56)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}
\KeywordTok{names}\NormalTok{(blood14)[}\KeywordTok{names}\NormalTok{(blood14)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}
\KeywordTok{names}\NormalTok{(food78)[}\KeywordTok{names}\NormalTok{(food78)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}
\KeywordTok{names}\NormalTok{(food56)[}\KeywordTok{names}\NormalTok{(food56)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}
\KeywordTok{names}\NormalTok{(food14)[}\KeywordTok{names}\NormalTok{(food14)}\OperatorTok{==}\StringTok{"seriali"}\NormalTok{] <-}\StringTok{ "ID"}

@@

\NormalTok{NDAYS14 <-}\StringTok{ }\NormalTok{blood14 }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(ID, Ndays)}

\NormalTok{NDAYS56 <-}\StringTok{ }\NormalTok{blood56 }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(ID, Ndays)}

\NormalTok{NDAYS78 <-}\StringTok{ }\NormalTok{blood78 }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(ID, NDays) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{rename}\NormalTok{(}\DataTypeTok{Ndays =}\NormalTok{ NDays)}

\NormalTok{NDAYS <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(NDAYS14, NDAYS56, NDAYS78)}
\NormalTok{NDAYS}\OperatorTok{$}\NormalTok{ID <-}\StringTok{ }\KeywordTok{as.numeric}\NormalTok{(NDAYS}\OperatorTok{$}\NormalTok{ID)}

@@

\NormalTok{## dta ready to be analysed in STATA}
 \CommentTok{# change the path accordingly}
\KeywordTok{write_dta}\NormalTok{(CW3CB3_7regss, }\StringTok{"../CW3CB3_7regss.dta"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

**********************************************************
// variables need to be log transformed                 //
**********************************************************

### 2018-08-18 paper/Chapters/Chapter4.tex
limitation 1

Surprisingly, low carbohydrate eaters whose dietary recordings suggested that they were mostly following a regular temporal meals pattern turned out to consume the highest amount of total energy among the three types of carbohydrate eaters. Detailed profiling of energy composition according to the time slots revealed that low carbohydrate eaters actually had higher proportion of energy contributed by both alcohol and fat. High percentage of fat consumption was shown in all 7 time slots, energy coming from alcohol exceeded more than one fifth of the total energy after 8 pm. These findings explained why they were actually consuming the highest energy among the three types of carbohydrate eaters. However, we also found that participants consuming low carbohydrate food had higher prevalence of diabetes, hypertension, and obesity. These health issues might possibly lead them (or advices were given from their physicians) to replace carbohydrates in their diet to other energy sources such as fat, protein, or even alcohol. Therefore, there is a possibility that they chose to follow low carbohydrate diets out of health purpose, but many of them were replacing carbohydrates with higher energy condensed food or even alcohol at night which might indeed be a public health concern.

Next, when looking into the details of the timing and composition of the energy intake among the moderate carbohydrate eaters, we realised that although these individuals did not consume as much alcohol as low carbohydrate eaters at night, they consumed the highest amount of energy, especially during time period as late as after 10 pm. People fell into moderate carbohydrate eaters group seemed to have the tendency of having their food or meals later than the other two types of carbohydrate eaters. They consumed the highest amount of carbohydrates and also total energy among three types of carbohydrate eaters during the following time slots: 9-12 noon, 2-5 pm, 8-10 pm, and 10pm-6am. These individuals in the NDNS RP were younger, mostly single, with lower average income, and lower education level. They might possibly correspond to the "late eaters" defined by previous studies \parencite{leech2017temporal, mansukhani2018investigating}.

Lastly, the high carbohydrate eaters identified by our MLCA models had the highest absolute total amount of carbohydrate intake. Most of their energy intake occured during 6-9 am, 12-2 pm, and 5-8 pm time slots while their average energy consumption was not the highest (between low and moderate carbohydrate eaters). High carbohydrate eaters were also found to be the ones that consumed the least amount of energy after 8 pm. Therefore, contrary to what was anticipated at the beginning, people who followed high percentage carbohydrate days for most of their time, were actually eating a healthier diet compared with the other two eating patterns.

@@

Among men, who were classified as moderate carbohydrate eaters probably had lower odds of having hypertension, after adjustment of age, live with partner or not, educational level, BMI, smoking status and total energy intake. As discussed above, moderate carbohydrate eaters turned to have meals (or energy consumption) later in time compared with high carbohydrate eaters, but low carbohydrate eaters also consumed large amount of energy (from both fat and alcohol) at night. Therefore, considering that moderate carbohydrate eaters were younger than low carbohydrate eaters (although age was adjusted in the full models), there is probably reverse causality exists here (also due to the nature of cross-sectional study). That is, they were potentially both late eaters, however, with their age increased (and so as increased health-related problems/concern) some of them modified their habits, such as quit smoking, replace carbohydrate food with other energy sources which lead to the phenomenon of lower odds of hypertension in moderate carbohydrate eaters. Although the hypothesis cannot be determined by NDNS RP's cross-sectional data, if these theories were true, the energy sources they used to replace carbohydrate in their diet were apparently not very wisely chosen.

Among women, whether living with a partner became an interaction factor for the associations between carbohydrate eating patterns and BMI and abdominal obesity (WC). Directions of the associations were opposite to each other depending on whether women were living with a partner or not. This interaction effect was more obvious when looking at abdominal obesity measurement. High carbohydrate eaters had lower BMI and WC in those lived with their partner, while moderate carbohydrate eaters had higher WC in those who lived alone after adjustment of age, education level, smoking status, total energy intake, and alcohol consumption. High carbohydrate eaters who were characterised with high and early in the day energy consumption and low fat and alcohol intake may reflect a healthier diet and lifestyle, but this might be different between women who lived alone and those who lived with their partners. It was often assumed that live alone may associated with lower diversity of food intake, and a higher likelihood of having an unhealthy dietary pattern \parencite{hanna2015relationship}. Therefore, there may be differences in the actual contents consumed in the high carbohydrates eaters, or there may be other social, psychological or lifestyle related factors related with living alone which we did not included in the models, so that the inverse association between high carbohydrate eating pattern and BMI or abdominal obesity were only observed among women who lived with partners, further analysis is needed. Whereas the reason why moderate carbohydrate eaters' WC was larger than low carbohydrate eaters only among women who lived alone is unknown, given that the evidence of this association was weak and borderline significant, whether it was just a false positive result should be explored in other studies.

@@

\section{Strengths and limitations}

\begin{itemize}
	\item MLCA ignored the order of observation days.
	\item We used the maximum probability rule and ignored that these are just probabilities. but inclusive MLCA might perform better.
	\item we do not have the information of whether the participants were doing job that requires shift work.
	\item dietary report under estimation
	\item findings may not be generalizable to populations from other countries.
\end{itemize}

>>>>

Surprisingly, low carbohydrate eaters whose dietary recordings suggested that they were mostly following a regular temporal meals pattern turned out to be consuming the highest amount of total energy among the three types of carbohydrate eaters. Detailed profiling of energy composition according to the time slots revealed that low carbohydrate eaters actually had higher proportion of energy contributed by both alcohol and fat. High percentage of fat consumption was shown in all 7 time slots, energy coming from alcohol exceeded more than one fifth of the total energy after 8 pm. These findings explained why they were actually consuming the highest energy among the three types of carbohydrate eaters. However, we also found that participants consuming low carbohydrate food had higher prevalence of diabetes, hypertension, and obesity. These health issues might possibly lead them (or advices were given from their physicians) to replace carbohydrates in their diet to other energy sources such as fat, protein, or even alcohol. Therefore, there is a possibility that they chose to follow low carbohydrate diets out of health purpose, but many of them were replacing carbohydrates with higher energy condensed food or even alcohol at night which might indeed be a public health concern.

Next, when looking into the details of the timing and composition of the energy intake among the moderate carbohydrate eaters, we realised that although these individuals did not consume as much alcohol as low carbohydrate eaters did at night, they consumed the highest amount of energy, especially during time period as late as after 10 pm. People fell into moderate carbohydrate eaters group seemed to have the tendency of having their food or meals later than high carbohydrate eaters. They consumed the highest amount of carbohydrates and also total energy among three types of carbohydrate eaters during the following time slots: 9-12 noon, 2-5 pm, 8-10 pm, and 10pm-6am. These individuals in the NDNS RP were younger, largely single, with lower average income, and lower education level. They might possibly correspond to the "late eaters" defined by previous studies \parencite{leech2017temporal, mansukhani2018investigating}.

Finally, the high carbohydrate eaters identified by our MLCA models had the highest absolute total amount of carbohydrate intake. Most (nearly 70\%) of their total energy intake occurred during 6-9 am, 12-2 pm, and 5-8 pm time slots although their total energy consumption was not the highest (between low and moderate carbohydrate eaters). High carbohydrate eaters were also found to be the ones that consumed the least amount of energy after 8 pm. Therefore, contrary to what was anticipated at the beginning, people who followed high percentage carbohydrate days for most of their time, were actually eating a healthier diet compared with the other two eating patterns.

@@

Among men, who were classified as moderate carbohydrate eaters probably had lower odds of having hypertension, after adjustment of age, live with partner or not, educational level, BMI, smoking status and total energy intake. As discussed above, moderate carbohydrate eaters turned to have meals (or energy consumption) later in time compared with high carbohydrate eaters, but low carbohydrate eaters also consumed large amount of energy (from both fat and alcohol) at night. Therefore, considering that moderate carbohydrate eaters were younger than low carbohydrate eaters (although age was adjusted in the full models), there is probably reverse causality exists here (also due to the nature of cross-sectional study). That is, they were potentially both late eaters, however, with increased age (and so as increased health-related problems/concern) some of them modified their habits, such as quit smoking, replace carbohydrate food with other energy sources which lead to the phenomenon of lower odds of hypertension in moderate carbohydrate eaters. Although these hypotheses cannot be determined by the cross-sectional data from NDNS RP, if they were true, the energy sources they used to replace carbohydrate in their diet were apparently not very wisely chosen.

Among women, whether living with a partner became an interaction factor for the associations between carbohydrate eating patterns and BMI and abdominal obesity (WC). Directions of the associations were opposite to each other depending on whether women were living with a partner or not. This interaction effect was more obvious when looking at abdominal obesity measurement. High carbohydrate eaters had lower BMI and WC in those lived with their partner, while moderate carbohydrate eaters had higher WC in those who lived alone after adjustment of age, education level, smoking status, total energy intake, and alcohol consumption. High carbohydrate eaters who were characterised with high and early in the day energy consumption and low fat and alcohol intake may reflect a healthier diet and lifestyle, but this might be different between women who lived alone and those who lived with their partners. It was often assumed that live alone may associated with lower diversity of food intake, and a higher likelihood of having an unhealthy dietary pattern \parencite{hanna2015relationship}. Therefore, there may be differences in the actual contents consumed in high carbohydrates eating women, or there may be other social, psychological or lifestyle related factors related with living alone which were not measured or did not included in the models, so that the inverse association between high carbohydrate eating pattern and BMI or abdominal obesity were only observed among women who lived with partners, further analysis is needed. Whereas the reason why moderate carbohydrate eaters' WC was larger than low carbohydrate eaters only among women who lived alone is unknown, given that the evidence of this association was weak and borderline significant, whether it was just a false positive finding should be explored in other studies.

@@

\section{Limitations and strengths}

There are several limitations in the current project that merits consideration.

Firstly, we ignored the order of observation days in the MLCA models. The food consumption diaries were accomplished by participants for at least 3 out of 4 consecutive days. In the multilevel analyses, these 3 or 4 days' observations were treated as if they were exchangeable in the models. Since one's diet might change according to season, day of the week (weekdays or weekends), or sometimes depends on what they had consumed the day before, exchangeability of their diaries is a strong assumption and cannot be overcome by the MLCA models adopted here. Other statistical techniques, which could take the order or the longitudinal nature of the data into account, such as repeated measures latent class analysis (RMLCA), latent transition analysis (LTA) \parencite{collins2010latent}, or latent class growth analysis (LCGA) \parencite{davidian2008growth,jung2008introduction,andruff2009latent} are not applicable for the NDNS RP dataset. For the purpose of maintaining the response rate, flexibility was allowed for the NDNS RP participants in their choice of which day to begin their diary.  Additionally, these models require that the 3 or 4 repeated measurements of food diaries should be recorded at the same time points longitudinally. LCGA models (also called growth mixture modelling) will need to model the change of the odds of the probability over time as a function (quadratic, or cubic) which is apparently not the objective in the current project.

Secondly, the classification of individuals to latent carbohydrate eating classes are defined by maximum posterior probability assignment rule. This approach assigns individuals to the class for which they have the highest posterior probability of membership \parencite{nagin2005group}.

\begin{itemize}
	\item We used the maximum probability rule and ignored that these are just probabilities. but inclusive MLCA might perform better.
	\item we do not have the information of whether the participants were doing job that requires shift work.
	\item dietary report under estimation
	\item findings may not be generalizable to populations from other countries.
\end{itemize}
